### 2014-09-11 paper/paper.Rmd
A few small fixes here and there

Reducing inequities in education outcomes is of great interest to politicians. Dissecting the causes of inequality is a much tougher challenge with no panacea. Nonetheless, tackling inequality in educational outcomes is essential from both moral and economic viewpoints. The Programme for International Student Assessment (PISA) is a triennial survey conducted by the Organization for Economic Cooperation and Development (OECD) with a rotating additional emphasis on one of mathematics, reading, and science. In 2012, the additional emphasis was on mathematics. All 34 member countries of the OECD and 31 partnering countries and economies participated in the survey. This represents over 80% of the global economy. The OECD estimates that if all students could reach a level-2 proficiency in mathematics -- a level-2 proficiency means that a student can only handle the ``simplest and most obvious tasks" -- it would add a collective US$200 trillion to the world's collective GDP.

@@

The data was made available by the OECD as part of a data challenge for the useR! 2014 conference. Entries to the competition can be found at http://beta.icm.edu.pl/PISAcontest/. We learned a lot from this data, and won one of the prizes. This is a description of the seven main things we learned about 15 year olds, their math, science and reading skills, and the effects of other factors.

@@

Although, not shown here, it also should be noted that this is the difference between the average scores for boys and girls. There is a slightly different story for each country if we look at the top math score for each geneder only, where we would notice that the top score in the USA is achieved by a girl, and is substantially higher than the highest boys score. There is far more individual variability than the overall average differences.

@@

The school table has information about shortages on items viewed as being important for good instruction. The plot below examines at three of these, lack of teachers, lack of computers and shortage of library materials, in relation to math, reading and science scores from the student table, across countries. For most countries scores decrease on average the greater the shortage.

@@

For each country we have provided loess smoothed curves of the relationship between mathematics and the time spent studying outside of school. The most prominent feature that we observe is the decreasing benefit of studying. For every country there is evidence of a substantial positive relationship for the students who spend less time studying. However, those students who continue to study for long hours see diminishing returns for their efforts. It is possible that there may be a selection issue occuring where more talented students spend less time studying as they complete their homework faster. However, the relationship shows those spending

@@

# 5. Parents matter!
### a. Number of parents in the home
Looking at the number of parents in a household shows that student in two-parent household are on average performing better than student's not in a two-parent household. Students with just a mother at home do tend to perform better than students with only a father at home. Students with no parents at home are performing the worst, on average.

@@

### b. Parent's Occupation Status
One of the variables the survey addressed was the occupation status of each of the parents. There were five possible categories: “Full-Time”, “Part-Time”, “Not working, but looking for a job”, “Other (e.g. home duties, retired)”, and no response. The distribution of the number of parents who selected each of the categories can be seen in the bar chart. The chart shows that father’s primarily have full-time jobs. However, there are large numbers of mothers who work full-time, part-time, or are stay-at-home moms. With women making up a larger portion of the work-force, it is interesting to look at how the parent’s occupation relates to the student’s performance in mathematics, reading, and science.

@@

The Figure ?*? shows the average performance of student’s performance based on the parent’s occupational status. The countries are sorted according to the average score for students with a "stay-at-home" mother. The averages for each parent in each of the work status categories are distinguished by color. As you can see in each of the plots below, students’ with a mother who works part-time or full-time tend outperform those students with a stay-at-home mother. This pattern holds with only a few exceptions. Meanwhile, when the father is employed part-time students tend to have much lower achievement that if the father stayed at home. When the father has a full-time jobs, the student’s have the highest average achievement among the possible work status’ for the father. The patterns described hold across mathematics, reading, and science.

@@

Albania appears to be an outlier in many educational trends.  Many socioeconomic and school factors that play a role in other nations'test scores do not appear to affect Albania's test scores.  For example, Albania has one of the smallest gender gaps for math and reading scores.  Additionally number of possessions and truancy rates affected test scores for every nation except Albania.

@@

Generally students with more possessions are expected to perform better in math, reading and science.  The number of possessions is a stand in for a student's economic standing.  The survey records whether students have desks, computers, internet, books, a place to study and dozens of other possessions.  After aggregating total number of possessions together, the figure ?*? shows that increasing number of possessions is correlated with higher test scores in all nations except Albania.

@@

The PISA dataset contains hundreds of factors, and this article can describe just a few findings. The reader is welcome to explore the data themselves using an <a href="http://bit.ly/pisa2012_explorer"> interactive web tool</a>, made with <a href="http://www.rstudio.com">RStudio</a> and Shiny. Shiny is a open-source tool, developed by RStudio, that allows an investigator to create a web-based appication powered by R.

@@

Shiny, by RStudio

plyr, dplyr, and ggplot2, by Hadley Wickham

>>>>

Reducing inequities in education outcomes is of great interest to politicians and the public, alike. Dissecting the causes of inequality is a much tougher challenge with no panacea. Nonetheless, tackling inequality in educational outcomes is essential from both moral and economic viewpoints. The Programme for International Student Assessment (PISA) is a triennial survey conducted by the Organization for Economic Cooperation and Development (OECD) with a rotating additional emphasis on one of mathematics, reading, and science. In 2012, the additional emphasis was on mathematics. All 34 member countries of the OECD and 31 partnering countries and economies participated in the survey. This represents over 80% of the global economy. The OECD estimates that if all students could reach a level-2 proficiency in mathematics -- a level-2 proficiency means that a student can only handle the ``simplest and most obvious tasks" -- it would add US$200 trillion to the world's collective GDP.

@@

The data was made available by the OECD as part of a data challenge for the useR! 2014 conference. Entries to the competition can be found at http://beta.icm.edu.pl/PISAcontest/. We learned a lot from this data, and won one of the prizes. This article is a description of the seven main things we learned about 15-year olds, their math, science and reading skills, and some of the factors that affect these skills.

@@

Although, not shown here, it also should be noted that this is the difference between the average scores for boys and girls. There is a slightly different story for each country if we look at the top math score for each geneder only, where we would notice that the top score in the USA is achieved by a girl, and is substantially higher than the highest boys' score. There is far more individual variability than the overall average differences.

@@

Within the PISA dataset, the school table has information about shortages on items viewed as being important for good instruction. The plot below examines at three of these: lack of teachers, lack of computers and shortage of library materials. These factors are compared with math, reading and science scores from the student table, across countries. For most countries, scores decrease on average the greater the shortage.

@@

For each country we have provided loess-smoothed curves of the relationship between mathematics score and the time spent studying outside of school. The most prominent feature that we observe is the decreasing benefit of studying. For every country there is evidence of a substantial positive relationship for the students who spend less time studying. However, those students who continue to study for long hours see diminishing returns for their efforts. It is possible that there may be a selection issue occuring where more talented students spend less time studying as they complete their homework faster. However, the relationship shows those spending

@@

# 5. Parents matter!
### a. Number of parents in the home
Looking at the number of parents in a household shows that students in two-parent households are on average performing better than students not in a two-parent households. Students with just a mother at home do tend to perform better than students with only a father at home. Students with no parents at home are performing the worst, on average.

@@

### b. Parent's Occupation Status
One of the variables the survey addressed was the occupation status of each of the parents. There were five possible categories: “Full-Time”, “Part-Time”, “Not working, but looking for a job”, “Other (e.g. home duties, retired)”, and no response. The distribution of the number of parents who selected each of the categories can be seen in the bar chart. The chart shows that fathers primarily have full-time jobs. However, there are large numbers of mothers who work full-time, part-time, or are stay-at-home moms. With women making up a larger portion of the work-force, it is interesting to look at how the parents' occupations relate to students' performance in mathematics, reading, and science.

@@

The Figure ?*? shows the average performance of students' performance based on the parents' occupational statuses. The countries are sorted according to the average score for students with a "stay-at-home" mother. The averages for each parent in each of the work status categories are distinguished by color.  In each of the plots below, students with mothers who work part-time or full-time tend outperform those students with stay-at-home mothers. This pattern holds with only a few exceptions. Meanwhile, when the father is employed part-time, students tend to have much lower achievement that if the father stayed at home. Those students whose fathers have full-time jobs have the highest average achievement among the possible work statuses for the father. The patterns described hold across mathematics, reading, and science.

@@

Albania appears to be an outlier in many educational trends.  Many socioeconomic and school factors that play a role in other nations' test scores do not appear to affect Albania's test scores.  For example, Albania has one of the smallest gender gaps for math and reading scores.  Additionally number of possessions and truancy rates affected test scores for every nation except Albania.

@@

Generally, students with more possessions are expected to perform better in math, reading and science.  The number of possessions is a stand in for a student's economic standing.  The survey records whether students have desks, computers, internet, books, a place to study and dozens of other possessions.  After aggregating total number of possessions together, the figure ?*? shows that increasing number of possessions is correlated with higher test scores in all nations except Albania.

@@

The PISA dataset contains hundreds of factors, and this article can describe just a few findings. The reader is welcome to explore the data themselves using an  [interactive web tool](http://bit.ly/pisa2012_explorer), made with RStudio's and [Shiny](http://shiny.rstudio.com). Shiny is a open-source tool that allows an investigator to create web-based appications powered by R.

@@

shiny, plyr, dplyr, ggplot2
